---
title: "Emery_Assignment_Nonlinear_Temp_Night"
output:
  html_document:
    df_print: paged
---
Challange:

Fit monthly temperature response curves using a similar approach with the night data from harv (night).

NEE ~ a exp(bTA)

a is the base respiration rate when air temperature is 0 rC and b is an empirical coefficient.

Workflow:
1. Create a dataframe to store month parameter values (parms.Month).
2. Write a function to the fit model and extract parameters (nee.night).
3. Write a loop to fit monthly curves and add parameters to a dataframe (parms.Month).
4. Bootstrapping for error estimation.



```{r}
# Set your working directory to the Malone NLM Workshop
rm(list=ls())

load("C:/Users/Mere/Desktop/FIU Courses/Spring 2020/Quantative Ecology_WS/Nonlinear/NLM_Workshop.RData")

library(nlstools)
```

Temperature Response Curve

```{r}
plot( NEE ~ TA, data= night, xlab="Temperature")
```
```{r}
par(mai=c(1,1,0.1,0.1))
par(mfrow=c(2,2))
plot( NEE ~ TA, data= night)
plot( NEE ~ MONTH, data= night)
plot( PAR ~ MONTH, data= night)
plot( TA ~ MONTH, data= night)

jan <- subset(night, MONTH == 1)
feb <- subset(night, MONTH == 2)
march <- subset(night, MONTH == 3)
april <- subset(night, MONTH == 4)
may <- subset(night, MONTH == 5)
june <- subset(night, MONTH == 6)
july <- subset(night, MONTH == 7)
aug <- subset(night, MONTH == 8)
sept <- subset(night, MONTH == 9)
oct <- subset(night, MONTH == 10)
nov <- subset(night, MONTH == 11)
dec <- subset(night, MONTH == 12)

plot( NEE ~ TA, data= jan)
plot( NEE ~ TA, data= feb)
plot( NEE ~ TA, data= april)
plot( NEE ~ TA, data= may)
plot( NEE ~ TA, data= june)
plot( NEE ~ TA, data= july)
plot( NEE ~ TA, data= aug)
plot( NEE ~ TA, data= sept)
plot( NEE ~ TA, data= oct)
plot( NEE ~ TA, data= nov)
plot( NEE ~ TA, data= dec)


```

```{r}
##STARTING VALUES FOR NONLINEAR FUNCTIONS

# Selfstart for the trc:
trcModel <- function(TA, a, b) {
  y = a * exp(b * TA)
  return(y)
}
```


```{r}
# Create a function to find initial values for the selfstart function:

trc.int <- function (mCall, LHS, data){
  x <- data$TA
  y <- data$NEE
 
  a <-1.00703982 + -0.08089044* (min(na.omit(y)))
  b <- 0.051654 + 0.001400 * (min(na.omit(y)))
 
  value = list(a, b)
  names(value) <- mCall[c("a", "b")]
  return(value)
}
```


```{r}
# Selfstart function
SS.trc <- selfStart(model=trcModel,initial = trc.int)
```


```{r}
#Dataframe to store parms and se
# Create Dataframe to store the data:
parms.Month <- data.frame(
  MONTH=numeric(),
  a=numeric(),
  b=numeric(), 
  a.pvalue=numeric(),
  b.pvalue=numeric(), stringsAsFactors=FALSE, row.names=NULL)


parms.Month[1:12, 1] <- seq(1,12,1) # Creates time file to merge with parm file:

```


```{r}
#Functions: Use Intial Values in the model
nee.night <- function(dataframe){y.df = nls(NEE ~ a * exp(b*TA), 
                                            dataframe, start=list(a= iv$a , b=iv$b ),
                                            na.action=na.exclude, trace=F,
                                            control=nls.control(warnOnly=T))

y.df <- as.data.frame(cbind(t(coef(summary(y.df))[1:2, 1]), t(coef(summary(y.df)) [1:2, 4])))

names(y.df) <- c("a", "b", "a.pvalue", "b.pvalue")                      
return(y.df)}
```


```{r}
# This loop fits monthly models (1:12):
try(for(j in unique(night$MONTH)){
  print(j)
  
  iv <- getInitial(NEE ~ SS.trc('TA', "a", "b"), data = night[which(night$MONTH == j),]) 
  
  y4 <- try(nee.night(night[which(night$MONTH == j),]), silent=T) # Fit night model
  
  try(parms.Month[c(parms.Month$MONTH == j ), 2:5 ] <- cbind(y4), silent=T)
  
  rm(y4)
}, silent=T)

parms.Month
```
```{r}
# Create file to store parms and se
boot.NEE <- data.frame(parms.Month[, c("MONTH")]); names (boot.NEE) <- "MONTH"
boot.NEE$a.est<- 0
boot.NEE$b.est<- 0
boot.NEE$a.se<- 0
boot.NEE$b.se<- 0
```


```{r}
# Night Model:
for ( j in unique(boot.NEE$MONTH)){
  print(j)
  y1 <-night[which(night$MONTH == j),]
  
  iv <- getInitial(NEE ~ SS.trc('TA',"a", "b"), data = y1) 
  
  night.fit <- nls(NEE ~ a * exp(b*TA), 
                   data=y1, start=list(a= iv$a , b=iv$b ),
                   na.action=na.exclude, trace=F,
                   control=nls.control(warnOnly=T))
  
  results <- nlsBoot(night.fit, niter=100 )
  a <- t(results$estiboot)[1, 1:2]
  names(a) <- c('a.est', 'b.est')
  b <- t(results$estiboot)[2, 1:2]
  names(b) <- c('a.se', 'b.se')
  c <- t(data.frame(c(a,b)))
  boot.NEE[c(boot.NEE$MONTH == j), 2:5] <- c[1, 1:4]
  rm(night.fit, a, b, c, results, y1)
}
```


```{r}
trc <- merge( parms.Month, boot.NEE)

trc
trc$a<-round(trc$a,3)
trc$b<-round(trc$b,3)
trc$a.pvalue<-"p<0.001"
trc$b.pvalue<-round(trc$b.pvalue,2)
trc$b.pvalue[trc$b.pvalue==0]<-"p<0.001"
trc$a.est<-round(trc$a.est,3)
trc$b.est<-round(trc$b.est,2)
trc$a.se<-round(trc$a.se,3)
trc$b.se<-round(trc$b.se,3)
write.csv(trc,"C:/Users/Mere/Desktop/QEcology/Assignment_Workshop_2_Nonlinear-models_files\\Temp_Resp_Curve.csv",row.names = F)
```

```{r}
library(knitr)
library(kableExtra)
kable(trc[,1:5],caption="Table 1. Temperature Response Curve",results='asis')


```

