---
title: "Timeseries_Emery"
output: html_notebook
---
The primary objectives of the workshop:
1. Create a time series object.
2. Decompose a time series.
3. Test for stationarity and detecting autocorrelation.
4. Choose the order of an ARIMA model.
5. Use explanatory series to improve models.

Library and Data
```{r}
rm(list=ls())
library(zoo)
library(tseries)
library(forecast)
library(xts)
load("C:/Users/Mere/Desktop/Spring2020_QEcology/Assignment_Workshop_3_Timeseries/ARIMA_Workshop.RData")
```

Part 1: 
1. Create Timeseries objects:

```{r}
nee <- ts( mangroves$nee, start= 1, frequency=30)
```

Visualize data:
```{r}
par(mfrow=c(1,1), mai=c(0.25,0.8,0.1, 0.1))
plot( nee, typ="l", ylab= "NEE", xlab="Time", main="Net Ecological Exchange through Time")
```
```{r}
plot(nee)
lines(tsclean(nee), col="red")
```

```{r}
nee <- tsclean(nee)

```

2. Decompose the timeseries: 
```{r}
nee.d <- decompose(nee, 'multiplicative')
plot(nee.d)
```

3.a. Test for stationarity
```{r}
# p-value < 0.05 indicates the TS is stationary
adf.test(nee )

```

3.b. Detecting Autocorrelation:
```{r}
acf(nee, lag.max=45)
```

```{r}
pacf(nee, lag.max=45)
```

4. Fitting an ARIMA Model: 

```{r}
arima.nee1 <-auto.arima(nee, trace=TRUE)
```

Residual plots show a smaller error range, more or less centered around 0. We can observe that AIC is smaller for the second model structure as well:
```{r}
tsdisplay(residuals(arima.nee1), lag.max=45)
```
```{r}
par(mfrow=c(1,1))
plot(nee , typ="l",xlab="Time", ylab= "NEE", main= "Model 1: ARIMA NEE"); lines(fitted(arima.nee1),col="red")
```

```{r}
arima.nee2 <-arima(nee , order=c(10,1,3), seasonal= list(order=c(2,0,2))) #accounting for some of the outliers
```

```{r}
tsdisplay(residuals(arima.nee2), lag.max= 30)
```
```{r}
# You want to minimize AIC
AIC(arima.nee1, arima.nee2)
```
```{r}
par(mfrow=c(1,1))
plot(nee , typ="l"); lines(fitted(arima.nee2),col="red")
```

Test of independence at all lags
```{r}
# Measuring for significant difference from white noise.
# You need a p-value greater than 0.05!
checkresiduals(arima.nee2, lag=36)
```
```{r}
par(mfrow=c(1,1))
plot(nee , typ="l", xlab= "Time", ylab= "NEE", main="Model 2: ARIMA NEE"); lines(fitted(arima.nee2),col="red")
```

Forecast plot

```{r}
plot(forecast(arima.nee2, h=30),xlab = "Time", ylab = "NEE", main = "Forecasts for Model 2: NEE")
```

1. Create a timeseries object
```{r}
sal <- ts(mangroves$salinity.max, start= 1, frequency=30)
```

Visualize data:
```{r}
par(mfrow=c(1,1), mai=c(0.25,0.8,0.1, 0.1))
plot(sal , typ="l", ylab= "Salinity", xlab="")
```

You want to remove any outliers that could bias the model by skewing statistical summaries. 

```{r}
plot(sal , typ="l", ylab= "Salinity", xlab="")
lines(tsclean(sal) , col="red")
```

Clean the timeseries
```{r}
sal <- tsclean(sal)
```

2. Decompose the time series
```{r}
sal.d <- decompose(sal, 'multiplicative')
plot(sal.d)
```

3.a. Test for stationarity
```{r}
# p-value < 0.05 indicates the TS is stationary
adf.test(sal)
```

```{r}
adf.test(diff(sal))
```

4. Explore correlations.
```{r}
ccf( diff(sal),nee, na.action = na.pass, lag.max=40, plot=TRUE)

```

5. Explore Models of NEE
```{r}
arima.nee3 <-auto.arima(nee, xreg=c(diff(sal),0), trace=TRUE)
```

```{r}
par(mfrow=c(1,1))
plot(nee , typ="l",ylab="NEE",main="Model 3: Difference in Salinity"); lines(fitted(arima.nee3),col="red")
```


```{r}
#compare to current model
AIC(arima.nee2, arima.nee3 )
```
So, adding salinity to nee did not improve the model. Maybe extreme salinity is more important. Lets create
a salinity index to ID when salinity values are greater than 25n ppt.

```{r}
sal.i <- sal
sal.i[sal.i < 25 ]<- 0
sal.i[sal.i >= 25 ]<- 1
plot(sal.i)

```

Now try adding the extreme salinity indicator into the model to see if this is an improvement:

```{r}
arima.nee4 <-auto.arima(nee, xreg=sal.i, trace=TRUE)
```

```{r}
AIC(arima.nee2,arima.nee4 )

```

```{r}
checkresiduals(arima.nee4, lag=36)
```

```{r}
par(mfrow=c(1,1))
plot(nee , typ="l",ylab="NEE",main="Model 4: Extreme Salinity Indicator "); lines(fitted(arima.nee4),col="red")

```

Challenge:

Developing another model using: water.tmax
```{r}
water.tmax<-ts(mangroves$water.tmax,start= 1, frequency = 30)
```

Visualize data
```{r}
par(mfrow=c(1,1),maic(0.25,0.8,0.1,0.1))
plot(water.tmax, typ="l", ylab="Water_Temp_Max",xlab="")
```

```{r}
plot(water.tmax)
lines(tsclean(water.tmax),col="red")
```

```{r}
water.tmax<-tsclean(water.tmax)
```

2.Decompose the timeseries

```{r}
water.tmax.d<-decompose(water.tmax, 'multiplicative')
plot(water.tmax.d)
```
3.a. Test for stationarity

```{r}
adf.test(water.tmax)
```

```{r}
adf.test(diff(water.tmax))

diff_watert<-diff(water.tmax)
```

4. Explore Correlation (look for significant lags)
```{r}
ccf(diff_watert,nee, na.action=na.pass, lag.max=40, plot=TRUE)
```

5. Explore Models of NEE
```{r}
arima.nee5wtmax <-auto.arima(nee, xreg = c(diff_watert,0),trace=TRUE)
```

Comparing AIC
```{r}
AIC(arima.nee1,arima.nee2,arima.nee3,arima.nee4,arima.nee5wtmax)

```

Check Residuals
```{r}
checkresiduals(arima.nee5wtmax,lag=36)
```
```{r}
par(mfrow=c(1,1))
plot(nee , typ="l",ylab="NEE",main="Model 5: Difference in Max Water Temperature"); lines(fitted(arima.nee5wtmax),col="red")
```

Developing another model using: PAR
```{r}
PAR<-ts(mangroves$par,start= 1, frequency = 30)
```

Visualize data
```{r}
par(mfrow=c(1,1),maic(0.25,0.8,0.1,0.1))
plot(PAR, typ="l", ylab="PAR",xlab="")
```

```{r}
plot(PAR)
lines(tsclean(PAR),col="red")
```

```{r}
PAR<-tsclean(PAR)
```

2.Decompose the timeseries

```{r}
PAR.d<-decompose(PAR, 'multiplicative')
plot(PAR.d)
```
3.a. Test for stationarity

```{r}
adf.test(PAR)
```

```{r}
adf.test(diff(PAR))

diff_PAR<-diff(PAR)
```

4. Explore Correlation (look for significant lags)
```{r}
ccf(diff_PAR,nee, na.action=na.pass, lag.max=40, plot=TRUE)
```

5. Explore Models of NEE
```{r}
arima.nee6PAR <-auto.arima(nee, xreg = c(diff_PAR,0),trace=TRUE)
```

Comparing AIC
```{r}
AIC(arima.nee1,arima.nee2,arima.nee3,arima.nee4,arima.nee5wtmax,arima.nee6PAR)

```

Check Residuals
```{r}
checkresiduals(arima.nee6PAR,lag=36)
```
Nee with Diff PAR as the red line
```{r}
par(mfrow=c(1,1))
plot(nee, type="l",ylab="NEE",main="Model 6: Difference in PAR");lines(fitted(arima.nee6PAR),col="red")
```

Developing another model using: Air Temp
```{r}
tair<-ts(mangroves$tair,start= 1, frequency = 30)
```

Visualize data
```{r}
par(mfrow=c(1,1),maic(0.25,0.8,0.1,0.1))
plot(tair, typ="l", ylab="Air Temp",xlab="")
```

```{r}
plot(tair)
lines(tsclean(tair),col="red")
```

```{r}
tair<-tsclean(tair)
```

2.Decompose the timeseries

```{r}
tair.d<-decompose(tair, 'multiplicative')
plot(tair.d)
```
3.a. Test for stationarity

```{r}
adf.test(tair)
```

```{r}
adf.test(diff(tair))

diff_tair<-diff(tair)
```

4. Explore Correlation (look for significant lags)
```{r}
ccf(diff_tair,nee, na.action=na.pass, lag.max=40, plot=TRUE)
```

5. Explore Models of NEE
```{r}
arima.nee7tair<-auto.arima(nee, xreg = c(diff_tair,0),trace=TRUE)
```

Comparing AIC
```{r}
AIC(arima.nee1,arima.nee2,arima.nee3,arima.nee4,arima.nee5wtmax,arima.nee6PAR,arima.nee7tair)

```

Check Residuals
```{r}
checkresiduals(arima.nee7tair,lag=36)
```
Nee with Diff tair as the red line
```{r}
par(mfrow=c(1,1))
plot(nee, type="l", ylab="NEE",main="Model 7: Difference in Air Temp");lines(fitted(arima.nee7tair),col="red")
```





```{r}
plot(forecast(arima.nee7tair, h=30),xlab = "Time", ylab = "NEE", main = "Forecasts for Model 7: Difference in Air Temperature")
```


```{r}
mangrove.summary<-by(mangroves$nee, mangroves$salinity.max, mangroves$water.tmax, mangroves$par, mangroves$tair,summary, na.rm=TRUE)

mangrove.summary<-by(mangroves$nee, mangroves$salinity.max,summary, na.rm=TRUE)

mangrove.summary

var_mangroves<-sapply(mangroves[,c(2:5,7)],FUN="var")
var_mangroves

mean_mangroves<-sapply(mangroves[,c(2:5,7)],FUN="mean")
mean_mangroves

min_mangroves<-sapply(mangroves[,c(2:5,7)],FUN="min")
max_mangroves<-sapply(mangroves[,c(2:5,7)],FUN="max")
rbind(mean_mangroves,min_mangroves,max_mangroves,var_mangroves)

df_mangroves<-(data.frame(rbind(mean_mangroves,min_mangroves,max_mangroves,var_mangroves),row.names = c("Mean","Min","Max","Variance")))

df_mangroves

colnames(df_mangroves)[which(colnames(df_mangroves)=='nee')]<-'NEE'
colnames(df_mangroves)[which(colnames(df_mangroves)=='par')]<-'PAR'
colnames(df_mangroves)[which(colnames(df_mangroves)=='tair')]<-'Air Temperature'
colnames(df_mangroves)[which(colnames(df_mangroves)=='water.tmax')]<-'Max Water Temperature'
colnames(df_mangroves)[which(colnames(df_mangroves)=='salinity.max')]<-'Max Salinity'

df_mangroves

df_mangroves$NEE<-round(df_mangroves$NEE,4)
df_mangroves$PAR<-round(df_mangroves$PAR,4)
df_mangroves$`Air Temperature`<-round(df_mangroves$`Air Temperature`,4)
df_mangroves$`Max Water Temperature`<-round(df_mangroves$`Max Water Temperature`,4)
df_mangroves$`Max Salinity`<-round(df_mangroves$`Max Salinity`,4)

```

```{r}
write.csv(df_mangroves,"C:/Users/Mere/Desktop/Spring2020_QEcology/Assignment_Workshop_3_Timeseries\\Summary_mangroves_data.csv")
```

