---
title: "Species Distribution Model_Inclass"
arther: "Meredith Emery"
output: html_notebook
---


```{r}
rm(list=ls())
library("sp")
library("raster")
library("maptools")
library("rgdal")
library("dismo")

bioclim.data <- getData(name = "worldclim",
                        var = "bio",
                        res = 2.5,
                        path = "/Users/Mere/Desktop/FIU Courses/Spring 2020/Quantative Ecology_WS/Species Distribution Model")

```

Inport Data
```{r}
obs.data <- read.csv(file = "/Users/Mere/Desktop/Spring2020_QEcology/Assignment_Workshop_4_Species_Distribution_Model/SDM_Data.csv")
obs.data<-na.omit(obs.data)

summary(obs.data)

```


```{r}
# Notice NAs - drop them before proceeding
obs.data <- obs.data[!is.na(obs.data$latitude), ]

# Make sure those NA's went away
summary(obs.data)
```

```{r}
# Determine geographic extent of our data
max.lat <- ceiling(max(obs.data$latitude))
min.lat <- floor(min(obs.data$latitude))
max.lon <- ceiling(max(obs.data$longitude))
min.lon <- floor(min(obs.data$longitude))
geographic.extent <- extent(x = c(min.lon, max.lon, min.lat, max.lat))
```

```{r}
# Load the data to use for our base map
data(wrld_simpl)

# Plot the base map
plot(wrld_simpl, 
     xlim = c(min.lon, max.lon),
     ylim = c(min.lat, max.lat),
     axes = TRUE, 
     col = "grey95", main="Individual Observation Locations")

# Add the points for individual observation
points(x = obs.data$longitude, 
       y = obs.data$latitude, 
       col = "olivedrab", 
       pch = 20, 
       cex = 0.75)
# And draw a little box around the graph
box()
```


Building a model and visualizing results

```{r}
# Crop bioclim data to geographic extent of saguaro
bioclim.data <- crop(x = bioclim.data, y = geographic.extent)

# Build species distribution model
bc.model <- bioclim(x = bioclim.data, p = obs.data) #problem with the data format that bioclim wont run. 

```

Fixing the format problem

```{r}
# Drop unused column
obs.data <- obs.data[, c("latitude", "longitude")]

# Build species distribution model
bc.model <- bioclim(x = bioclim.data, p = obs.data)

```

```{r}
head(obs.data)
```


Reverse the columns (x,y) so that is (longitude, latitiude)
```{r}
# Reverse order of columns
obs.data <- obs.data[, c("longitude", "latitude")]

# Build species distribution model
bc.model <- bioclim(x = bioclim.data, p = obs.data)
```

Predict presence from model
```{r}
predict.presence <- dismo::predict(object = bc.model, x = bioclim.data, ext = geographic.extent)
```

Plot

```{r}
# Plot base map
plot(wrld_simpl, 
     xlim = c(min.lon, max.lon),
     ylim = c(min.lat, max.lat),
     axes = TRUE, 
     col = "grey95",main="Plot2")

# Add model probabilities
plot(predict.presence, add = TRUE)

# Redraw those country borders
plot(wrld_simpl, add = TRUE, border = "grey5")

# Add original observations
points(obs.data$longitude, obs.data$latitude, col = "olivedrab", pch = 20, cex = 0.75)
box()
```


The pseudo-absence point

```{r}
# Use the bioclim data files for sampling resolution
bil.files <- list.files(path = "/Users/Mere/Desktop/FIU Courses/Spring 2020/Quantative Ecology_WS/Species Distribution Model/wc2-5", 
                          pattern = "*.bil$", 
                          full.names = TRUE)

# We only need one file, so use the first one in the list of .bil files
mask <- raster(bil.files[1])

# Randomly sample points (same number as our observed points)
background <- randomPoints(mask = mask,     # Provides resolution of sampling points
                   n = nrow(obs.data),      # Number of random points
                   ext = geographic.extent, # Spatially restricts sampling
                   extf = 1.25)             # Expands sampling a little bit
```

```{r}
head(background)
```

We can also visualize them on the map, like we did for the observed points:
```{r}
# Plot the base map
plot(wrld_simpl, 
     xlim = c(min.lon, max.lon),
     ylim = c(min.lat, max.lat),
     axes = TRUE, 
     col = "grey95",
     main = "Presence and pseudo-absence points")

# Add the background points
points(background, col = "grey30", pch = 1, cex = 0.75)

# Add the observations
points(x = obs.data$longitude, 
       y = obs.data$latitude, 
       col = "olivedrab", 
       pch = 20, 
       cex = 0.75)

box()
```

Now that we have our pseudo-absence points, we need to take one more step. Getting a more traditional-range-map-looking figure requires post hoc evaluation of the model. To do this evaluation, we are going to build the model using only part of our data (the training data), reserving a portion of the data for evaluation of the model after it is build (the testing data). We are going to reserve 20% of the data for testing, so we use the kfold function in the dismo package to evenly assign each observation to a random group.

```{r}
# Arbitrarily assign group 1 as the testing data group
testing.group <- 1

# Create vector of group memberships
group.presence <- kfold(x = obs.data, k = 5) # kfold is in dismo package

```

Now pause for a minute and take a look at that group.presence vector we just created:

```{r}
head(group.presence)
```

```{r}
# Should see even representation in each group
table(group.presence)
```

The output of table shows how many points have been assigned to each of the five groups. In this case, we can see that the points have been evenly distributed, with 20% of the points in group 1, our testing group.

We use the group.presence vector with the observed data to separate our observations into a training data set and a testing data set:

```{r}
# Separate observations into training and testing groups
presence.train <- obs.data[group.presence != testing.group, ]
presence.test <- obs.data[group.presence == testing.group, ]

# Repeat the process for pseudo-absence points
group.background <- kfold(x = background, k = 5)
background.train <- background[group.background != testing.group, ]
background.test <- background[group.background == testing.group, ]

```


Training and testing

Now that we have (1) our pseudo-absence points and (2) separate training and testing data, we can re-build the model, evaluate its performance, and draw a more aesthetically pleasing map. We build the model with the bioclim function as before, but instead of using all the observations in obs.data we only use the training data stored in presence.train:

```{r}
# Build a model using training data
bc.model <- bioclim(x = bioclim.data, p = presence.train)

# Predict presence from model (same as previously, but with the update model)
predict.presence <- dismo::predict(object = bc.model, 
                                   x = bioclim.data, 
                                   ext = geographic.extent)
```


We now take that model, and evaluate it using the observation data and the pseudo-absence points we reserved for model testing. We then use this test to establish a cutoff of occurrence probability to determine the boundaries of the saguaro range.

```{r}
# Use testing data for model evaluation
bc.eval <- evaluate(p = presence.test,   # The presence testing data
                    a = background.test, # The absence testing data
                    model = bc.model,    # The model we are evaluating
                    x = bioclim.data)    # Climatic variables for use by model

# Determine minimum threshold for "presence"
bc.threshold <- threshold(x = bc.eval, stat = "spec_sens")

```

The threshold function offers a number of means of determining the threshold cutoff through the stat parameter. Here we chose "spec_sens", which sets “the threshold at which the sum of the sensitivity (true positive rate) and specificity (true negative rate) is highest.” For more information, check out the documentation for threshold (?threshold, remember?).

And finally, we can use that threshold to paint a map with the predicted range of the saguaro!

```{r}
# Plot base map
plot(wrld_simpl, 
     xlim = c(min.lon, max.lon),
     ylim = c(min.lat, max.lat),
     axes = TRUE, 
     col = "grey95", main="Predicted range of the saguaro")

# Only plot areas where probability of occurrence is greater than the threshold
plot(predict.presence > bc.threshold, 
     add = TRUE, 
     legend = FALSE, 
     col = "olivedrab")

# And add those observations
points(x = obs.data$longitude, 
       y = obs.data$latitude, 
       col = "black",
       pch = "+", 
       cex = 0.75)

# Redraw those country borders
plot(wrld_simpl, add = TRUE, border = "grey5")
box()
```

Plot above does not look right.. t plotted a large portion of the map green. Let’s look at what we actually asked R to plot, that is, we plot the value of predict.presence > bc.threshold. So what is that?

```{r}
predict.presence > bc.threshold
```

The comparison of these two rasters produces another raster with values of only 0 or 1: 0 where the comparison evaluates as FALSE (i.e., when the value in a grid cell of predict.presence is less than or equal to the value in the corresponding grid cell of bc.threshold) and 1 where the comparison evaluates at TRUE. Since there are two values in this comparison (the 0 and 1 in the values field), we need to update what we pass to the col parameter in our plot call. Instead of just passing a single value, we provide a color for 0 (NA) and a color for 1 ("olivedrab"):

```{r}
# Plot base map
plot(wrld_simpl, 
     xlim = c(min.lon, max.lon),
     ylim = c(min.lat, max.lat),
     axes = TRUE, 
     col = "grey95", main="Range of Saguaro cactus using the Bioclim Model")

# Only plot areas where probability of occurrence is greater than the threshold
plot(predict.presence > bc.threshold, 
     add = TRUE, 
     legend = FALSE, 
     col = c(NA, "olivedrab"))

# And add those observations
points(x = obs.data$longitude, 
       y = obs.data$latitude, 
       col = "black",
       pch = "+", 
       cex = 0.75)

# Redraw those country borders
plot(wrld_simpl, add = TRUE, border = "grey5")
box()
```

A final note on our approach: the map we have drawn presents a categorical classification of whether a particular point on the landscape will be suitable or not for the species of interest. This classification relies quite heavily on the value of the threshold (see bc.threshold and the documentation for threshold) and the pseudo-absence points. Given that we used random sampling to generate those pseudo-absence points, there is potential for variation in the predicted range if you run this code more than once (try it! if you re-run the code from the point of creating the pseudo-absence points, you are almost guaranteed a different map.). There are a number of approaches to dealing with this variation, and the paper by Barbet-Massin et al. (2012) is a great resource. I’ll leave it as homework for you to determine which approach is most appropriate here!

```{r}
write.csv(bioclim.data,"C:/Users/Mere/Desktop/FIU Courses/Spring 2020/Quantative Ecology_WS/Species Distribution Model\\bioclim.data.csv")


summary(bioclim.data)

bioclim.data_2<-na.omit(bioclim.data)
summary(bioclim.data_2)

```

```{r}

?`raster-package`
m <-cellStats(bioclim.data, stat='mean')
m

```

