---
title: "Quantative Ecoloty Project: Lake Annie Nutrient Timeseries"
output: html_notebook
---

 

```{r}
rm(list=ls())
library(zoo)
library(tseries)
library(forecast)
library(xts)

```

ARIMA TIMESERIES with 2006-2018 monthly data
```{r}
Secchi_M=read.csv("C:/Users/Mere/Desktop/WS_QEcology_2020/Project/Data Used/LA_2006_2018_Secchi_Nutrients.csv")


Secchi_M$Date <- as.Date(paste(Secchi_M$Month, Secchi_M$Day, Secchi_M$Year, sep="/"), format="%m/%d/%Y")
class(Secchi_M$Date)


Secchi_M_ts<-ts(Secchi_M$Secchi_m,start=1,frequency = 12)
par(mfrow=c(1,1), mai=c(0.25,0.8,0.1, 0.1))
plot(Secchi_M$Date, Secchi_M_ts, typ="l",ylim=rev(range(Secchi_M_ts)), ylab="Secchi Depth (meters)",xlab="Date")

```

```{r}
plot(Secchi_M_ts, ylim=rev(range(Secchi_M_ts)))
lines(tsclean(Secchi_M_ts),col="red")

```
```{r}
Secchi_M_tsc<- tsclean(Secchi_M_ts)

Secchi_M_d<- decompose(Secchi_M_tsc,'multiplicative')
plot(Secchi_M_d)
```

Test for Stationarity
```{r}
adf.test(Secchi_M_tsc)

# p-value < 0.05 indicates the TS is stationary
```
```{r}
adf.test(diff(Secchi_M_tsc))
```

Detecting Autocorrelation

```{r}
acf(diff(Secchi_M_tsc))
```
```{r}
pacf(diff(Secchi_M_tsc))
```


Fitting an ARIMA MODEL
```{r}
arima_Secchi_1_M<-auto.arima(diff(Secchi_M_tsc),trace = TRUE)
```


```{r}
arima_Secchi_1_M<-auto.arima(diff(Secchi_M_tsc),trace = TRUE)
tsdisplay(residuals(arima_Secchi_1_M))

```


```{r}
Secchi_M_diff<--diff(Secchi_M_tsc)

arima_Secchi_2_M<-arima(Secchi_M_diff , order=c(4,0,1), seasonal= list(order=c(1,0,0)))

arima_Secchi_2_M
```

```{r}
tsdisplay(residuals(arima_Secchi_2_M))
```


```{r}
arima_Secchi_3_M<-arima(Secchi_M_diff , order=c(1,4,1), seasonal= list(order=c(1,0,0)))

tsdisplay(residuals(arima_Secchi_3_M))
```

```{r}
arima_Secchi_4_M<-arima(Secchi_M_diff , order=c(1,0,4), seasonal= list(order=c(1,0,0)))

tsdisplay(residuals(arima_Secchi_4_M))
```
```{r}
par(mfrow=c(1,1))

plot(Secchi_M_diff,type="l"); lines(fitted(arima_Secchi_4_M),col="red")
```

```{r}
AIC(arima_Secchi_1_M,arima_Secchi_2_M,arima_Secchi_3_M,arima_Secchi_4_M)
```

Test for independence. Ljung-Box

```{r}
checkresiduals(arima_Secchi_4_M)

# Measuring for significant difference from white noise.
# You need a p-value greater than 0.05!
```



Forcasts
```{r}
par(mfrow=c(1,1))
plot(forecast(arima_Secchi_4_M, h=24))

```

Explanatory variables

```{r}
DOC_M<-ts(Secchi_M$DOC_ppm,start = 1,frequency = 12)


par(mfrow=c(1,1), mai=c(0.25,0.8,0.1, 0.1))

plot(Secchi_M$Date,DOC_M,typ="l",ylab="DOC",xlab="Date")

```

```{r}
plot(DOC_M,typ="l",ylab="DOC",xlab="Date")
lines(tsclean(DOC_M),col="red")
```

```{r}
DOC_M_tsc<-tsclean(DOC_M)
```

Decompose the time series

```{r}
DOC_d <- decompose(DOC_M_tsc, 'multiplicative')
plot(DOC_d)
```

Test for Stationarity

```{r}
adf.test(DOC_M_tsc)
```

```{r}
DOC_diff<--diff(DOC_M_tsc)

adf.test(DOC_diff)
```

Explore correlations
Look for significant lags

```{r}

ccf(DOC_diff,Secchi_M_diff,na.action = na.pass, plot = TRUE)

```

```{r}
arima_DOC <-auto.arima(Secchi_M_diff, xreg=c(DOC_diff), trace=TRUE)
```

```{r}
AIC(arima_Secchi_1_M,arima_Secchi_2_M,arima_Secchi_3_M,arima_Secchi_4_M,arima_DOC)
```

```{r}
checkresiduals(arima_DOC)
```

```{r}
Date_diff<-Secchi_M$Date[c(-1,0)]
par(mfrow=c(1,1), mai=c(0.25,0.8,0.1, 0.1))
plot(Secchi_M_diff,typ="l", ylab="Secchi Depth (meters)",xlab="Year");lines(fitted(arima_DOC),col="red")
```
```{r}
plot(Secchi_M$Date, Secchi_M_ts, typ="l",ylim=rev(range(Secchi_M_ts)), ylab="Secchi (meters)",xlab="Date")
```



Chlorophyl

```{r}
Chloro_M<-ts(Secchi_M$Chl_a_ug_per_L,start = 1,frequency = 12)


par(mfrow=c(1,1), mai=c(0.25,0.8,0.1, 0.1))

plot(Secchi_M$Date,Chloro_M,typ="l",ylab="Chlorophyll a (ug/L)",xlab="Date")

```

```{r}
plot(Chloro_M,typ="l",ylab="Chlorophyll a",xlab="Date")
lines(tsclean(Chloro_M),col="red")
```

```{r}
Chloro_M_tsc<-tsclean(Chloro_M)
```

Decompose the time series

```{r}
Chloro_d <- decompose(Chloro_M_tsc, 'multiplicative')
plot(Chloro_d)
```

Test for Stationarity

```{r}
adf.test(Chloro_M_tsc)
```

```{r}
Chloro_diff<--diff(Chloro_M_tsc)

adf.test(Chloro_diff)
```

Explore correlations
Look for significant lags

```{r}

ccf(Chloro_diff,Secchi_M_diff,na.action = na.pass, plot = TRUE)

```

```{r}
arima_Chloro <-auto.arima(Secchi_M_diff, xreg=c(Chloro_diff), trace=TRUE)
```

```{r}
AIC(arima_Secchi_1_M,arima_Secchi_2_M,arima_Secchi_3_M,arima_Secchi_4_M,arima_DOC,arima_Chloro)
```

```{r}
checkresiduals(arima_Chloro)
```

```{r}
Date_diff<-Secchi_M$Date[c(-1,0)]
par(mfrow=c(1,1), mai=c(0.25,0.8,0.1, 0.1))
plot(Secchi_M_diff,typ="l", ylab="Secchi (meters)",xlab="Year");lines(fitted(arima_Chloro),col="red")
```

Ground water Stage

```{r}
GWS_M<-ts(Secchi_M$Groundwater_Stage_m_amsl,start = 1,frequency = 12)


par(mfrow=c(1,1), mai=c(0.25,0.8,0.1, 0.1))

plot(Secchi_M$Date,GWS_M,typ="l",ylab="DOC",xlab="Date")

```

```{r}
plot(GWS_M,typ="l",ylab="Groundwater Stage m amsl",xlab="Date")
lines(tsclean(GWS_M),col="red")
```

```{r}
GWS_M_tsc<-tsclean(GWS_M)
```

Decompose the time series

```{r}
GWS_d <- decompose(GWS_M_tsc, 'multiplicative')
plot(GWS_d)
```

Test for Stationarity

```{r}
adf.test(GWS_M_tsc)
```

```{r}
GWS_diff<--diff(GWS_M_tsc)

adf.test(GWS_diff)
```

Explore correlations
Look for significant lags

```{r}

ccf(GWS_diff,Secchi_M_diff,na.action = na.pass, plot = TRUE)

```

```{r}
arima_GWS <-auto.arima(Secchi_M_diff, xreg=c(GWS_diff), trace=TRUE)
```

```{r}
AIC(arima_Secchi_1_M,arima_Secchi_2_M,arima_Secchi_3_M,arima_Secchi_4_M,arima_DOC,arima_Chloro,arima_GWS)
```

```{r}
checkresiduals(arima_GWS)
```

```{r}
Date_diff<-Secchi_M$Date[c(-1,0)]
par(mfrow=c(1,1), mai=c(0.25,0.8,0.1, 0.1))
plot(Secchi_M_diff,typ="l", ylab="Secchi (meters)",xlab="Year");lines(fitted(arima_GWS),col="red")
```


Total Phosphorus

```{r}
TP_M<-ts(Secchi_M$TP_ppm,start = 1,frequency = 12)


par(mfrow=c(1,1), mai=c(0.25,0.8,0.1, 0.1))

plot(Secchi_M$Date,TP_M,typ="l",ylab="Total Phosphorus",xlab="Date")

```

```{r}
plot(TP_M,typ="l",ylab="TP",xlab="Date")
lines(tsclean(TP_M),col="red")
```

```{r}
DOC_M_tsc<-tsclean(DOC_M)
```

Decompose the time series

```{r}
TP_d <- decompose(TP_M, 'multiplicative')
plot(TP_d)
```

Test for Stationarity

```{r}
adf.test(TP_M)
```

```{r}
TP_diff<--diff(TP_M)

adf.test(TP_diff)
```

Explore correlations
Look for significant lags

```{r}

ccf(TP_diff,Secchi_M_diff,na.action = na.pass, plot = TRUE)

```

```{r}
arima_TP <-auto.arima(Secchi_M_diff, xreg=c(TP_diff), trace=TRUE)
```

```{r}
AIC(arima_Secchi_1_M,arima_Secchi_2_M,arima_Secchi_3_M,arima_Secchi_4_M,arima_DOC, arima_Chloro, arima_GWS, arima_TP)
```

```{r}
checkresiduals(arima_TP)
```

```{r}
Date_diff<-Secchi_M$Date[c(-1,0)]
par(mfrow=c(1,1), mai=c(0.25,0.8,0.1, 0.1))
plot(Secchi_M_diff,typ="l", ylab="Secchi (meters)",xlab="Year");lines(fitted(arima_TP),col="red")
```

Total Nitrogen
```{r}
TN_M<-ts(Secchi_M$TN_ppm,start = 1,frequency = 12)


par(mfrow=c(1,1), mai=c(0.25,0.8,0.1, 0.1))

plot(Secchi_M$Date,TN_M,typ="l",ylab="Total Nitrogen",xlab="Date")

```

```{r}
plot(TN_M,typ="l",ylab="TN",xlab="Date")
lines(tsclean(TN_M),col="red")
```

```{r}
TN_M_tsc<-tsclean(TN_M)
```

Decompose the time series

```{r}
TN_d <- decompose(TN_M_tsc, 'multiplicative')
plot(TN_d)
```

Test for Stationarity

```{r}
adf.test(TN_M_tsc)
```

```{r}
TN_diff<--diff(TN_M_tsc) 

adf.test(TN_diff)
```

Explore correlations
Look for significant lags

```{r}

ccf(TN_diff,Secchi_M_diff,na.action = na.pass, plot = TRUE)

```

```{r}
arima_TN <-auto.arima(Secchi_M_diff, xreg=c(TN_diff), trace=TRUE)
```

```{r}
AIC(arima_Secchi_1_M,arima_Secchi_2_M,arima_Secchi_3_M,arima_Secchi_4_M,arima_DOC, arima_Chloro, arima_GWS, arima_TP, arima_TN)
```

```{r}
checkresiduals(arima_TN)
```

```{r}
Date_diff<-Secchi_M$Date[c(-1,0)]
par(mfrow=c(1,1), mai=c(0.25,0.8,0.1, 0.1))
plot(Secchi_M_diff,typ="l", ylab="Secchi Depth (meters)",xlab="Year");lines(fitted(arima_TN),col="red")
```

Summary of ARIMA DOC

```{r}

summary(arima_DOC)

arima_DOC

```



Summary Table of Variables

```{r}
Secchi_M=read.csv("C:/Users/Mere/Desktop/WS_QEcology_2020/Project/Data Used/LA_2006_2018_Secchi_Nutrients.csv")

head(Secchi_M)

var_LA<-sapply(Secchi_M[,c(4:5,7:10)],FUN="var")
var_LA

mean_LA<-sapply(Secchi_M[,c(4:5,7:10)],FUN="mean")
mean_LA

min_LA<-sapply(Secchi_M[,c(4:5,7:10)],FUN="min")
max_LA<-sapply(Secchi_M[,c(4:5,7:10)],FUN="max")
rbind(mean_LA,min_LA,max_LA,var_LA)

df_LA_1<-(data.frame(rbind(mean_LA,min_LA,max_LA,var_LA),row.names = c("Mean","Min","Max","Variance")))

df_LA_1

colnames(df_LA)[which(colnames(df_LA)=='Secchi_m')]<-'Secchi Depth (meters)'
colnames(df_LA)[which(colnames(df_LA)=='DOC_ppm')]<-'DOC (ppm)'
colnames(df_LA)[which(colnames(df_LA)=='Groundwater_Stage_m_amsl')]<-'Ground Level (meters above mean seal level)'
colnames(df_LA)[which(colnames(df_LA)=='TP_ppm')]<-'TP (ppm)'
colnames(df_LA)[which(colnames(df_LA)=='TN_ppm')]<-'TN (ppm)'

df_LA_1

df_LA_1$`TN_ppm` <-round(df_LA_1$`TN_ppm`,4)
df_LA_1$`TP_ppm`<-round(df_LA_1$`TP_ppm`,4)
df_LA_1$`DOC_ppm`<-round(df_LA_1$`DOC_ppm`,4)
df_LA_1$`Secchi_m`<-round(df_LA_1$`Secchi_m`,4)
df_LA_1$`Groundwater_Stage_m_amsl`<-round(df_LA_1$`Groundwater_Stage_m_amsl`,4)
df_LA_1$`Chl_a_ug_per_L`<-round(df_LA_1$`Chl_a_ug_per_L`,4)

```

```{r}
write.csv(df_LA_1,"C:/Users/Mere/Desktop/WS_QEcology_2020/Project\\Summary_LA_2006_2018_data_2.csv")
```

